A pipeline to analyse time-course gene expression data

### 2019-11-13 scripts/manuscript.Rmd
Small edits + fixes onto the manuscript

The `moanin_model` object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used, the basis, and the degrees of freedom of the model.

@@

@@

Since we arbitrarily chose a subset of genes based on our filter, we would
like all genes to have a chance of assignment to a cluster, as well as a score
for whether it is a good assignment.  The scoring and label step allows us to
assign genes to clusters that were removed during the filtering step, yet are
good matches to the centroids found during the clustering.

@@

```{r}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = splines_kmeans_score_and_label(
    data, kmeans_clusters)
labels = scores_and_labels$labels

# Let's keep only the list of genes that have a label.
labels = unlist(labels[!is.na(labels)])
cluster2Score = names(labels[labels==cluster_of_interest])
```

**Nelle: This is currently for debugging purposes. The distribution of scores
seems very strange to me, with a lot of genes that don't match at all to any
clusters.**

```{r visualize_scores_all_figure, fig.height=6, fix.width=6}
# Get the best score for each gene
scores = apply(scores_and_labels$scores, 1, min)
hist(scores, breaks=(1:50/50), col="black", main="Distribution of scores")
```

**Compared to the EPICON project, a lot of genes seem to have very very poor
scores. The threshod on the EPICON project on the score was around 0.55, while
here, it is close to 0.87. The scores range from 0 to 1, with 1 being no
correlation between the gene and the centroid.**

```{r visualize_scores, fig.height=6, fix.width=6}
# Get the best score for each gene
par(mfrow=c(4, 5))
scores = apply(scores_and_labels$scores, 1, min)
labels = apply(scores_and_labels$scores, 1, which.min)
for(cluster_id in 1:n_clusters){
    hist(scores[labels == cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""))
}

labels = unlist(labels[!is.na(labels)])

@@

Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in%
cluster2Score)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes

```{r fig.height=6, fig.width=6}
# order them by their score
cluster2Score = cluster2Score[order(scores_and_labels$scores[cluster2Score, cluster_of_interest],
				    decreasing = TRUE)]
dataToPlot = rbind(data[cluster2Score[1:4], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
plot_splines_data(dataToPlot, moanin_model,
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

>>>>

The `moanin_model` object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used, the basis matrix, and the degrees of freedom of the model.

@@

The basis matrix contains the value of all of the spline basis function for
each sample of the moment. As such, replicate samples will have the same
values.

```{r basis_matrix}
kable(head(moanin_model$basis))
```

@@

Since we arbitrarily chose a subset of genes based on our filter, we would
like all genes to have a chance of assignment to a cluster, as well as a score
for whether it is a good assignment.  The scoring and label step allows us to
assign genes to clusters that were removed during the filtering step, yet are
good matches to the centroids found during the clustering. The score
corresponds to a goodness-of-fit between each gene and each cluster, computed
as follows:

$$
S(y_i, \mu_k) =  \frac{1}{S_0(k)} \times \underset{a_{iG_j}, b_{iG_j}}{\text{min}}\sum_{j}
  {\left(b_{iG_j}y_{ij} + a_{iG_j}
   - \mu_k(t_j)\right)}^2 \,,
$$

where $\mu_k$ is the centroid of cluster $k$ and $y_i$ the gene of interest.
The scoring function thus returns a value between 0 and 1, 0 being the best
score possible and 1 the worst score possible (no correlation between the gene
and the cluster centroid).

@@

```{r assigning_scores_and_labels}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = splines_kmeans_score_and_label(
    data, kmeans_clusters)
labels = scores_and_labels$labels

# Let's keep only the list of genes that have a label.
scores = rowMin(scores_and_labels$scores[!is.na(labels), ])
labels = unlist(labels[!is.na(labels)])
genes_in_cluster2 = names(labels[labels==cluster_of_interest])
```

Let us now, for each cluster, visualize the distribution of scores of the
genes assigned to that cluster.

```{r visualize_scores, fig.height=6, fix.width=6}
# Get the best score for each gene
par(mfrow=c(4, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(scores[labels == cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
}

@@

Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in%
genes_in_cluster2)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes

```{r fig.height=6, fig.width=6}
# order them by their score
cluster2Score = cluster2Score[order(scores_and_labels$scores[genes_in_cluster2, cluster_of_interest])]
dataToPlot = rbind(
    data[cluster2Score[1:4], ],
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
plot_splines_data(dataToPlot, moanin_model,
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

**NV: this were the worst scoring genes. I fixed this and the plots now look
much better**
